---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# genderOlympics

<!-- badges: start -->
<!-- badges: end -->

The `genderOlympics` package provides tools and visualisations to explore gender participation trends in the Olympic Games, spanning from 1900 to 2016 for both Summer and Winter Olympics. This package includes a comprehensive dataset, `olympics_participation`, with detailed information about each athlete's participation, and an interactive Shiny app for exploring trends across years, sports, and seasons.

For more detailed documentation, visit the [pkgdown site](https://etc5523-2024.github.io/assignment-4-packages-and-shiny-apps-Billy0624/).

## Installation

To install the development version of `genderOlympics` from GitHub, use the following command. Note that you'll need the remotes package for this installation.

``` r
# If you don't have the remotes package installed, install it first
install.packages("remotes")

# Install the genderOlympics package from GitHub
remotes::install_github("ETC5523-2024/assignment-4-packages-and-shiny-apps-Billy0624")
```

Once installed, load the package with:

```{r}
library(genderOlympics)
```

## Dataset overview

The primary dataset included in this package, `olympics_participation`, contains information on over 270,000 athletes, covering:

  - Athlete demographics: name, age, sex, height, and weight

  - Event details: sport, event, medal won (if any), and team

  - Olympic metadata: year, season (Summer or Winter), city, and National Olympic Committee (NOC) code

Example of exploring the dataset:

```{r message=FALSE, warning=FALSE, error=FALSE}
library(dplyr)
library(ggplot2)

# Filter for athletics sport in the Summer Olympics
athletics_data <- olympics_participation |>
  filter(sport == "Athletics", season == "Summer") |>
  group_by(year, sex) |>
  summarise(count = n(), .groups = "drop")

# Plot the number of male and female athletes per year for athletics
ggplot(athletics_data, aes(x = year, y = count, color = sex)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Number of male and female athletes in Athletics over time",
    x = "Year",
    y = "Number of athletes",
    color = "Gender"
  ) +
  theme_minimal()
```

## Launching the Shiny App

The `genderOlympics` package includes a Shiny app for interactively exploring gender trends in the Olympics. The app allows users to:

  - Select the season (Summer or Winter).
  
  - Filter by a specific year range.

  - Choose specific sports or select all sports at once.

To launch the Shiny app, use:

```{r eval=FALSE}
run_app()
```

This app provides two main visualizations:

  - Gender participation over time: Displays the number of male and female athletes over time.

  - Sports participation by gender: Shows the number of sports in which male and female athletes participated over time.
  
# Contributing

If you'd like to contribute to `genderOlympics`, feel free to submit issues or pull requests on the GitHub repository. Contributions are always welcome!

For further details, see the [pkgdown site](https://etc5523-2024.github.io/assignment-4-packages-and-shiny-apps-Billy0624/).
