---
title: "Introduction to genderOlympics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to genderOlympics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Installation

You can install the `genderOlympics` package directly from GitHub using the remotes package. First, ensure you have the remotes package installed, then run the following command to install the package:

```{r eval=FALSE}
# If you don't have the remotes package installed, install it first
install.packages("remotes")

# Install the genderOlympics package from GitHub
remotes::install_github("ETC5523-2024/assignment-4-packages-and-shiny-apps-Billy0624")
```

Once installed, you can load the package as usual:

```{r setup}
library(genderOlympics)
```

The `genderOlympics` package allows users to explore gender trends in Olympic participation data, focusing on both Summer and Winter Olympics between 1900 and 2016. It comes with a dataset that includes details about each athlete's participation, including the year, event, sport, medal won (if any), and their country of representation.

Additionally, the package offers an interactive Shiny application that allows users to visually explore this data by filtering seasons, year ranges, and specific sports.

This vignette will guide you through the following:

  1. Loading the `olympics_participation` dataset.

  2. Using the provided functions, including launching the interactive Shiny app.

  3. Example use cases for filtering and analyzing the dataset.
    
# The olympics_participation dataset

The main dataset provided by this package is `olympics_participation`, which contains data for over 270,000 athletes who participated in the Olympics from 1900 to 2016.

```{r}
dplyr::glimpse(olympics_participation)
```

The dataset contains the following columns:

  - `id`: A unique identifier for each athlete.
  - `name`: The name of the athlete.
  - `sex`: Gender of the athlete ("M" for male, "F" for female).
  - `age`: Age of the athlete during the Olympics.
  - `height`: Athlete’s height (in centimeters).
  - `weight`: Athlete’s weight (in kilograms).
  - `team`: The country or team the athlete represented.
  - `noc`: National Olympic Committee code.
  - `games`: The specific Olympic Games the athlete participated in (e.g., "1992 Summer").
  - `year`: The year of the Olympic Games.
  - `season`: "Summer" or "Winter".
  - `city`: The host city of the Olympics.
  - `sport`: The sport in which the athlete participated.
  - `event`: The specific event within the sport.
  - `medal`: The medal the athlete won, if any ("Gold", "Silver", "Bronze", or NA if no medal was won).

## Example: Filtering the data

You can easily filter the dataset to answer specific questions about Olympic participation. For example, to get a list of all male basketball players from the 1992 Summer Olympics:

```{r}
basketball_men_1992 <- olympics_participation |>
  dplyr::filter(sport == "Basketball", sex == "M", year == 1992, season == "Summer")

head(basketball_men_1992)
```

## Data summary

You can also use basic summary functions to get a quick overview of the data. For example, the number of athletes who have won medals:

```{r}
table(olympics_participation$medal)
```

# Launching the Shiny App

The package comes with a built-in Shiny application that allows users to explore gender participation trends in the Olympics interactively.

To launch the app, use the following command:

```{r eval=FALSE}
run_app()
```

The Shiny app provides two visualisations:

  1. Gender participation over time: A plot that shows the number of male and female athletes who participated in the Olympics over time.

  2. Sports participation by gender: A plot that shows how many sports male and female athletes participated in over time.

## Shiny App Features

  - Season selection: Users can choose between Summer or Winter Olympics.

  - Year range filter: A slider allows users to filter the data by a range of years.

  - Sport selection: Users can select specific sports or all sports to visualise participation trends.

## Example: Using the Shiny App

After launching the Shiny app, users can interact with the data by choosing the season, filtering by year range, and selecting specific sports.

Here’s a typical usage flow:

  1. Select the season: Choose between "Summer" and "Winter".

  2. Adjust the year range: Use the slider to filter the data for a specific range of years (e.g., 1980 to 2016).

  3. Choose a sport: Select one or more sports from the dropdown list. Alternatively, check "Select all sports" to view data for all sports.

The app will generate two plots:

  - A line chart showing the number of male and female athletes for the selected sport(s) and season over time.

  - A chart showing how many sports male and female athletes participated in over time.
    
# Conclusion

The `genderOlympics` package provides an easy way to explore historical Olympic participation data, with a focus on gender trends. You can use the dataset for your own analysis or launch the Shiny app for a more interactive exploration.

For more details, refer to the help pages for each function and dataset:

```{r}
?olympics_participation
```

I hope this package helps you gain insights into gender representation in Olympic sports over time.
